---
title: 'Level 1 Data Cleaning: Clean the mtcars Dataset'
---

# Objective

The objective of this assignment is to practice cleaning and transforming a messy dataset using tidyverse functions. You will use skills like renaming and reordering columns, sorting rows, changing data types, mutating data, and using the stringr and forcats packages.

In this Level 1 Cleaning assignment, you will work with a simple dataset and focus on basic data cleaning tasks. Most tasks are outlined in the assignment script. You may want to review the [Data Cleaning Walkthrough](https://github.com/nrdowling/d2mr-assessment/tree/main/01_data-cleaning/00_cleaning-walkthrough) before you begin.

You may additionally or alternatively complete the [Level 2 Data Cleaning assignment](https://github.com/nrdowling/d2mr-assessment/tree/main/01_data-cleaning/01_cleaning-level-2). In Level 2, you will work with a more complex dataset and perform additional cleaning tasks with less direct instruction. The Level 2 assignment has more opportunities to demonstrating meeting course standards than this Level 1 assignment and is recommended for those who are already comfortable with the tasks in this assignment.

# Instructions

1. If you have not already done so, pull the latest changes from the `d2mr-assessment` repository to ensure you have the most up-to-date version of the assignment files. Confirm you are working in your fork of the repository.
2. Open `cleaning-level-1.qmd` in RStudio and follow the instructions in the Setup section below to load and inspect the (clean) `goal.mtcars` dataset. 
    - **Important:** The provided code makes a small modification to the original `mtcars` dataset to create a `goal.mtcars` dataset. You will use this goal dataset as a reference to clean the `messy-mtcars.csv` dataset, not the true original.
3. Follow the tasks described in the assignment script to clean the dataset and return it to its original state.
4. At several points in this document you will come across questions or non-coding exercises. Answer these questions in the text of this .qmd document, immediately below the question.
5. *Optional:* Continue to follow the instructions in the assignment script to clean the dataset above and beyond matching the original. 


### Tasks

**Reminder:** Your goal is to match `goal.mtcars` (created below), *not* the default `mtcars`.[^1]

[^1]: Why? The `mtcars` dataset includes models names as *row names* rather than as a variable/column. For the purposes of this cleaning exercise, it's more useful to treat model names as a variable.

1. **Renaming Columns:** Rename columns to match the original `mtcars` dataset column names.
2. **Reordering Columns:** Ensure the columns are in the correct order.
3. **Sorting Rows:** Sort the rows by `mpg` (miles per gallon) and then by `cyl` (number of cylinders).
4. **Changing Data Types:** Ensure `gear` and `carb` are factors and other columns are numeric.
5. **Mutating Data:** Create a new column `hp_per_cyl` calculated as `hp` (horsepower) divided by `cyl`.
6. **String Manipulation:** Use `stringr` to clean any unwanted spaces or characters in the `model` column.
7. **Handling Factors:** Use `forcats` to ensure the levels of `gear` are ordered as `3`, `4`, `5`.

# Setup

## Loading libraries and set seed

```{r}
#| label: setup
library(tidyverse)
set.seed(1234)
```


## Read in and inspect messy data

Read in and inspect the messy dataset `messy-mtcars.csv`.

```{r}

#| label: read-messy-data

### LEAVE THIS CHUNK AS-IS ###

# You *might* need to edit the filepath, but don't change anything else!

# Read in messy-mtcars.csv
messy.mtcars <- read_csv(
  ########################################
  "messy-mtcars.csv", ## <-- THIS IS THE ONLY THING IN THIS CHUNK YOU CAN CHANGE IF NECESSARY
  ########################################
  trim_ws = FALSE, name_repair = "minimal", col_types = cols(.default = col_character()))

# Inspect the data
head(messy.mtcars)

```

## Inspect the original mtcars dataset

```{r}
#| label: inspect-original-data

### LEAVE THIS CHUNK AS-IS ###

# Load the original mtcars dataset
data(mtcars)

# Create the "goal.mtcars" dataset
# Convert row names to a column called "model" (see note above about row names)
goal.mtcars <- mtcars %>%
  rownames_to_column(var = "model")

# Inspect the goal.mtcars dataset
head(goal.mtcars)

# Optionally inspect the original mtcars dataset to see what the row names vs column issue looks like
# head(mtcars)

```


QUESTIONS:

1. What are the differences between the messy dataset and the original mtcars dataset?

<!-- answer below -->
- Columns are out of order
- Column names don't match
- White space in the ' Car Model' column in the messy dataset
- Data types are incorrect 
- Numbers aren't rounded to the same number of decimal places
- Two extra columns in the messy dataset


2. What are the main issues you need to address in cleaning?

<!-- answer below -->
- Rename columns
- Reorder columns
- Change data types: columns to numeric
- Round numbers to the same number of decimal places 
- Remove two extra columns



# Clean the Dataset

## Creat "clean" dataset

```{r}

#| label: make-cleaning-dataset

# Create a dataset to work with during the cleaning process called "clean.mtcars"
clean.mtcars <- messy.mtcars

```


## Clean columns/variables


```{r}

#| label: rename-columns

# Rename columns to match the goal.mtcars dataset with rename()

clean.mtcars <- messy.mtcars %>%
    # The format is rename(new_name = old_name)
  rename(model = ' Car Model', mpg = 'mpg (Miles/Gallon)', cyl = cylinders, disp = dsip, hp = horsepower, drat = drat, wt = wt, qsec = qsec, vs = vs, am = am, gear = gearbox, carb = carburetors)

colnames(clean.mtcars)

```

```{r}

#| label: reorder-columns

# Reorder columns to match the goal.mtcars dataset
clean.mtcars <- clean.mtcars %>%
  relocate(model, mpg, cyl, disp, hp, drat, wt, qsec, vs, am, gear, carb)

colnames(clean.mtcars)

```

```{r}

#| label: correct-data-types

# convert to numeric 
clean.mtcars <- clean.mtcars %>% 
 mutate(
        mpg = as.numeric(mpg),
        cyl = as.numeric(cyl),
        disp = as.numeric(disp),
        hp = as.numeric(hp),
        drat = as.numeric(drat),
        wt = as.numeric(wt),
        qsec = as.numeric(qsec),
        vs = as.numeric(vs),
        am = as.numeric(am),
        gear = as.numeric(gear),
        carb = as.numeric(carb)
    ) 

```

## Checkpoint 1

```{r}

#| label: checkpoint-1

# Inspect the current state of the dataset

# Use all.equal() to check if clean.mtcars matches goal.mtcars

all.equal(clean.mtcars, goal.mtcars)
```

Questions:

3. The current state of the dataset does not yet match the `goal.mtcars` dataset. Explain the issues that `all.equal()` flags.  Are there groups of issues that can be addressed together?

<!-- answer below -->
- Attribute mismatch: overall class of the dataset differs
- Two extra columns in clean.mtcars
- Typos and inconsistet capitalization in the 'model' columns
- cly and hp have missing values that are not yet handled in the clean.mtcars dataset
- Rounding inconsistencies

3. Install (if necessary) and load the `daff` package. In the code chunk below, use the `diff_data()` function to identify the differences between `clean.mtcars` and `goal.mtcars`. 

```{r}
#| label: diff-data

# Install and/or load the daff package if needed
# install.package("daff")
library(daff)

# Use render_diff() and diff_data() to identify differences between clean.mtcars and goal.mtcars
render_diff(diff_data(clean.mtcars, goal.mtcars))

```

How is this method of identifying data discrepancies different from `all.equal()`? Which do you find more helpful at this point?

<!-- answer below -->
`diff_data()` tells us that there are differences between the two datasets. Since right now the two datasets are still very different, it tells us that there are differences basically in every row, which is not helpful in this case because I know it is very different. It is more helpful to use 'all.equal()' to clearly show me what kind of issues/discrepancies should be fixed.  

## Clean data values

Identifying specific value problems and cleaning them might take a little creativity at times! Use the troubleshooting/debugging workflows we've discussed to get as far as you can.

```{r}

#| label: remove-unwanted-spaces

# Remove unwanted whitespace in any columns required

clean.mtcars <- clean.mtcars %>%
    mutate(
        # Trim whitespace from the beginning and end of the string
        model = str_squish(model)
    )

```

```{r}

#| label: impute-missing-values

# Impute missing values for cyl and hp
clean.mtcars <- clean.mtcars %>%
  mutate(
    # Replace NA in `cyl` and 'hp' column with specific values
    cyl = ifelse(is.na(cyl) & model == 'CAdillac Fleetwood', 8, cyl), 
    cyl = ifelse(is.na(cyl) & model == 'FerrAri Dino', 6, cyl),   
    hp = ifelse(is.na(hp) & model == 'Merc 230', 95, hp),   
    hp = ifelse(is.na(hp) & model == 'Fiat 128', 66, hp),
    hp = ifelse(is.na(hp) & model == 'Porsche 914-2', 91, hp), 
  )
```

```{r}

#| label: round-numeric-values

# Round numeric values
clean.mtcars<- clean.mtcars %>% 
  # Round to the specific decimal places
    mutate(
        disp = round(disp, 1),
        drat = round(drat, 2),
        qsec = round(qsec, 2)
    )

all.equal(clean.mtcars, goal.mtcars)
render_diff(diff_data(clean.mtcars, goal.mtcars))
```

```{r}

#| label: fix-typos

# Fix typos in model names
## Hint: This can be completed with a 6-line mutate() statement

clean.mtcars <- clean.mtcars %>%
    mutate(model = str_replace_all(model, c(
        "SportAbout" = "Sportabout",
        "CAdillac" = "Cadillac",
        "ToyotA|T0yota" = "Toyota",  # Merged two replacements into one
        "PontiAc" = "Pontiac",
        "L0tus" = "Lotus",
        "FerrAri" = "Ferrari"
    )))

all.equal(clean.mtcars, goal.mtcars)
colnames(clean.mtcars)
print(clean.mtcars)

```

```{r}

#| label: change attributes

attributes(clean.mtcars)
attributes(goal.mtcars)

# Make both datasets tibbles
clean.mtcars <- as_tibble(clean.mtcars) 
goal.mtcars <- as_tibble(goal.mtcars)

```

```{r}

#| label: remove-extra-column

# Remove the extra column
clean.mtcars <- clean.mtcars %>%
  select(all_of(names(goal.mtcars)))

all.equal(clean.mtcars, goal.mtcars)

print(goal.mtcars)
print(clean.mtcars)

```


## Checkpoint 2

```{r}
#| label: checkpoint-2

# Inspect the dataset and compare to goal.mtcars

# all.equal()
all.equal(clean.mtcars, goal.mtcars)

# daff
render_diff(diff_data(clean.mtcars, goal.mtcars))

```

Questions:

5. Is your dataset identical to `goal.mtcars`? If not, what are the remaining issues? If there were any issues you could not resolve in code, describe a proposed solution in plain English.

<!-- answer below -->
The dataset is now identcal to 'goal.mtcars'.

# Optional Cleaning Tasks

You can call it a day here, or continue with the optional cleaning tasks below. These tasks are a chance to challenge yourself and practice your skills further. They are also additional opportunities to demonstrate mastery of course standards! 


## Additional guided cleaning

*Optional:* Complete any number of the following tasks to further clean the dataset. Each task should be in its own code chunk with an appropriate label and clearly commented code:

1. Create a new factor column `mpg_category` that categorizes `mpg` into "low", "medium", and "high" based on the distribution of `mpg` values.
```{r}
#| label: new factor column 'mpg_category' 

# Create a new factor column "mpg_category" based on mpg distribution
clean.mtcars <- clean.mtcars %>%
  mutate(mpg_category = case_when(
    mpg < quantile(mpg, 1/3) ~ "low",
    mpg < quantile(mpg, 2/3) ~ "medium",
    TRUE ~ "high"
  )) %>%
  mutate(mpg_category = factor(mpg_category, levels = c("low", "medium", "high")))

# Checkpoint
print(levels(factor(clean.mtcars$mpg_category)))
print(clean.mtcars)
```
2. Create a new factor column `wt_category` that categorizes `wt` into "light", "medium", and "heavy" based on the distribution of `wt` values, then reverse the order of the levels.
```{r}
#| label: new factor column 'wt_category' 

clean.mtcars <- clean.mtcars %>%
  mutate(wt_category = case_when(
    wt < quantile(wt, 1/3) ~ "light",
    wt < quantile(wt, 2/3) ~ "medium",
    TRUE ~ "heavy"
  ))

# Check the order before reversing it
print(levels(factor(clean.mtcars$wt_category)))

# Reverse the order of the levels
clean.mtcars <- clean.mtcars %>%
  mutate(wt_category = factor(wt_category, levels = c("light", "medium", "heavy"))) %>%
  mutate(wt_category = fct_rev(wt_category))

# Checkpoint
print(levels(factor(clean.mtcars$wt_category)))

```

3. Create a new column `mpg_per_cyl` that calculates `mpg` divided by `cyl`.
```{r}
#| label: new column 'mpg-per-cyl'

# Create the new column
clean.mtcars <- clean.mtcars %>%
  mutate(mpg_per_cyl = mpg / cyl)

# Checkpoint
print(clean.mtcars)

```

4. Filter the data to only include cars with automatic transmissions. (Hint: `?mtcars` will show you useful info.)
```{r}
#| label: filter data

#check useful info 
?mtcars

# Filter data for cars with automatic transmissions
auto.cars <- clean.mtcars %>%
  filter(am == 0)

# Checkpoint
head(auto.cars)

```



5. Identify which variable would be better suited as a logical variable, then convert it to logical.
```{r}
#| label: convert logical variable

#check useful info 
?mtcars

# Convert 'vs' to logical because it represents engine shape (0 and 1 values), which makes it a good candidate to convert to a logical variable (TRUE/FALSE)
clean.mtcars <- clean.mtcars %>%
  mutate(vs = as.logical(vs))

# Checkpoint
str(clean.mtcars$vs)

```



6. Sort the dataset by `mpg_category` then reverse alphabetically by `model` (so that models with the same `mpg_category` are sorted Z to A).
```{r}
#| label: sort by 'mpg_category'

# Sort by 'mpg_category', then reverse alphabetically by model
clean.mtcars <- clean.mtcars %>%
  arrange(mpg_category, desc(model))

# Checkpoint
head(clean.mtcars)
print(clean.mtcars)

```



7. Write the cleaned dataset as a csv file called `clean-mtcars.csv`, then read the csv back in maintaining the correct data types.
```{r}
#| label: write and read csv

# Write cleaned dataset to CSV
write_csv(clean.mtcars, "clean-mtcars.csv")

# Read back the dataset
clean.mtcars_read <- read_csv("clean-mtcars.csv", col_types = cols(
  mpg = col_double(),
  cyl = col_double(),
  disp = col_double(),
  hp = col_double(),
  drat = col_double(),
  wt = col_double(),
  qsec = col_double(),
  vs = col_logical(),
  am = col_double(),
  gear = col_double(),
  carb = col_double(),
  mpg_category = col_factor(levels = c("low", "medium", "high")),
  wt_category = col_factor(levels = c("light", "medium", "heavy"))
))

# Confirm correct types
str(clean.mtcars_read)

```




## Unguided cleaning and transformation

*Optional:* If you have the time and interest, continue transforming this dataset as you please. Create new columns based on the existing ones, reformat strings, try your hand at a regex replacement, summarize by groups (factor levels), visualize a simple relationship, or anything else you can think of. You can do this in addition to or instead of the suggested additional cleaning tasks above.


# Submission & Assessment

To submit:

1. Add & modify the `assessment.md` in this mini-project's directory:
    1. Check off all objectives you believe you have demonstrated
    2. Indicate which objectives you are meeting for the first time (if any)
    3. Complete any relevant open-ended items
2. Push your changes to your centralized assignment repository on GitHub. 
3. Confirm that Dr. Dowling and your section TA are added as collaborators to your repository.
4. Submit your work in your next open mini-project assignment by including the following information in the text box:
    1. The title of the assignment: "Level 1 Data Cleaning: Clean the mtcars Dataset"
    2. A link to the **directory** for this assignment in your centralized assignment repo


